---
title: "DataExploration_ggplot"
author:
- Douglas A. Campbell:
    institute: mta
    email: dcampbel@mta.ca
    correspondence: TRUE
institute:
  - mta: 'Department of Biology, Mount Allison University, 53 York St., Sackville NB, Canada, E4L 1C9.'
date: "`r format(Sys.Date())`"  
output:
  bookdown::html_document2:
    code_folding: show
    keep_md: yes
    fig_caption: yes
    toc: TRUE
    toc_float: TRUE
    toc_depth: 6
---

This is  [R Markdown](http://rmarkdown.rstudio.com) Notebook outlines approaches to using ggplot to explore data efficiently.

# Load Libraries
```{r load libraries}
library(tidyverse)
#by loading library(tidyverse) I get the dplyr, ggplot2 and purrr packages used in this tutorial
```

# Load Data
```{r load data}

Data <- readRDS(file.path("..", "data", "MultiCultiData.Rds"))
head(Data)

# Data <- Data %>%
#   select(
#     where(
#       ~!all(is.na(.x)) #remove columns of all na
#     )
#   ) %>%
#   slice_head(n = 120)
# 
# saveRDS(Data, file = file.path("..", "data", "MultiCultiData.Rds"))

```

# What Variables are available to explore?
```{r colnames}
colnames(Data)
```

# Step through plots of variables of interest using ggplot
Note I use the ggplot auto-assignment of X & Y axes labels and ranges.
For data exploration, do not hard code variable labels, nor axes ranges, to limit risks of missing data.
```{r ggplot interactively}
Data %>%
  ggplot() +
  geom_histogram(aes(lognormOD720_TimetoMax))

Data %>%
  ggplot() +
  geom_point(aes(x = OD720trunc_Lmax, y = OD720trunc_Lmu_corr))


```

With 125 variables for histograms of data distribution, or `r length(combn(125, 2))` combinations of data cross plots  is time consuming, and manually pasting many different variables into ggplot() calls is error prone.

# Set Axes variables External to ggplot
```{r plotting external variables}

#set fitting variables outside ggplot
XVar <- "OD720trunc_Lmax"
YVar <- "OD720trunc_Lmu_corr"
ColourVar <- "Strain"

#https://rlang.r-lib.org/reference/injection-operator.html
Data %>%
  ggplot()+
  geom_point(aes(x = !!sym(XVar), y = !!sym(YVar), colour = !!sym(ColourVar)))

```

# What about stepping through multiple variables for cross plots?
  Showing off, provoking thought
  https://aosmith.rbind.io/2018/08/20/automating-exploratory-plots/
  
```{r mapping ggplot variables}
#Identify response and explanatory variables for cross plotting
#In this case we will pick an arbitrary set of the first 10 variables for 'explanatory variables' and any variable that contains the string 'trunc' for 'response variables'.

ExplanVar <- names(Data)[1:10]
ResponseVar <- names(select(Data, contains('trunc')))

#set the 'names' of the values within the ExplanVar and ResponseVar character vectors; we 'name' each element of the vector with the element string itself
ExplanVar = set_names(ExplanVar)
ResponseVar = set_names(ResponseVar)

ExplanVar
#ResponseVar

Data %>%
  ggplot()+
  geom_point(aes(x = !!sym(XVar), y = !!sym(YVar), colour = !!sym(ColourVar)))

```
  
  